Multilevel spin models in cold atomic systems

### 2021-04-01 sun_phases/spin_model_paper/main.tex
updating multilevel spin model paper

% more shorthands
\newcommand{\up}{\uparrow}
\newcommand{\dn}{\downarrow}
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}
\newcommand{\g}{\text{g}}
\newcommand{\e}{\text{e}}
\renewcommand{\SS}{\mathbb{S}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\su}{\mathfrak{su}}
\newcommand{\MF}{\text{MF}}

\renewcommand{\ss}{\bar{\v s}\c\bar{\v s}}
\DeclareMathOperator{\diag}{diag}

@@

Here we review the realization of a collective SU($n$) spin model in a system of ultracold alkaline-earth(-like) atoms on an optical lattice.
Without external driving fields, the evolution of such atoms in their electronic ground state is governed by single-body kinetic and two-body interaction Hamiltonians of the form
\begin{align}
  H_{\t{kin}}
  &= -\f{J}{2} \sum_{\bk{j,j'},\mu} c_{j\mu}^\dag c_{j'\mu} + \t{h.c.}, \\
  H_{\t{int}}
  &= \f{U_0}{2} \sum_{j,\mu,\nu}
  c_{j\mu}^\dag c_{j\mu} c_{j\nu}^\dag c_{j\nu},
\end{align}
where $\bk{j,j'}$ denotes neighboring lattice sites $j$ and $j'$; $\mu,\nu\in\set{s,s-1,\cdots,-s}$ index orthogonal spin states of a spin-$s$ nucleus, with $s=\frac{n-1}{2}$ (e.g.~$s=\frac{9}{2}$ in the case of ${}^{87}$Sr with $10$ nuclear spin states); $c_{j\mu}$ is a fermionic annihilation operator, $J$ is a tunneling amplitude; and $U_0$ is a two-body on-site interaction energy.
In the present work, we neglect inter-site interactions and interaction-assisted hopping, which may become relevant in a sufficiently shallow lattice, namely when $J\gtrsim E_{\t{R}}$, where $E_{\t{R}}$ is the atomic lattice recoil energy.
For simplicity, we now assume a periodic lattice of $L$ sites, and expand the on-site fermionic operators in terms of operators addressing (quasi-)momentum modes $q$, $c_{j\mu}=\frac1{\sqrt{L}}\sum_k e^{-\i q\c j} c_{q\mu}$, finding that
\begin{align}
  H_{\t{kin}}
  &= -J\sum_q \cos\p{q} c_{q\mu}^\dag c_{q\mu}, \\
  H_{\t{int}}
  &= \f{U}{2N} \sum_{k,\ell,p,q,\mu,\nu}
  c_{k\mu}^\dag c_{\ell\mu} c_{p\nu}^\dag c_{q\nu}
  \times \delta_{k+p,\ell+q},
  \label{eq:H_int_momenta}
\end{align}
where $N$ is the total number of atoms on the lattice, we define $U\equiv U_0\times N/L$ for convenience, $\delta_{k+p,\ell+q}=1$ if $k+p=\ell+q$ and $0$ otherwise (enforcing conservation of momentum), and we work in units with lattice spacing $a=1$.

If the interaction energy $U_0$ is smaller than the single-particle bandwidth $\sim J$, then the mode-changing collisions in $H_{\t{int}}$ become off-resonant, motivating the frozen-mode approximation $\set{k,p}=\set{\ell,q}$ (which we assume for the remainder of this work).
The operator content of terms with $k=\ell$ and $p=q$ takes the form $c_{\ell\mu}^\dag c_{\ell\mu} c_{q\nu}^\dag c_{q\nu}$, so these terms contribute an overall energy $\frac12NU$ that can be neglected in the absence of coherent atom number fluctuations.
Defining the spin operators $s_{\mu\nu q}\equiv c_{q\mu}^\dag c_{q\nu}$, the remaining terms of the kinetic and interaction Hamiltonians become
\begin{align}
  H_{\t{kin}} &= -J\sum_q \cos\p{q} s_{\mu\mu q},
  \label{H_kin_start} \\
  H_{\t{int}} &= -\f{U}{2N} \sum_{p,q,\mu,\nu} s_{\mu\nu p} s_{\nu\mu q}.
  \label{eq:H_int_start}
\end{align}
Throughout this work, we will assume that atomic modes are singly-occupied, e.g.~due to the initialization of a spin-polarized state with one atom per lattice site, in which multiple occupation of an atomic mode is forbidden by Pauli exclusion.
In this case, we can neglect the fermionic origins of $H_{\t{int}}$ and instead treat our system as a bona fide spin model.

To further simplify notation and write the interaction Hamiltonian in a form suggestively similar to SU(2) spin alignment, we now construct the operator-valued spin matrix
\begin{align}
  \v s_q \equiv \sum_{\mu,\nu} s_{\mu\nu q} \op{\mu}{\nu},
\end{align}
and for any pair of such operator-valued matrices $\v A,\v B$, we define the inner product
\begin{align}
  \v A \c \v B \equiv \Tr_{\t{aux}}\p{\v A^\dag\v B}
  = \sum_{\mu,\nu} A_{\mu\nu}^\dag B_{\mu\nu},
\end{align}
where the trace is performed over the auxiliary matrix degrees of freedom.
These definitions allow us to write the spin Hamiltonian in Eq.~\eqref{eq:H_int_start} as
\begin{align}
  H_{\t{int}} = -\f{U}{2N} \sum_{p,q} \v s_p\c\v s_q
  = -\f{U}{2N}\v S\c\v S,
  &&
  \v S \equiv \sum_q \v s_q.
  \label{eq:H_int}
\end{align}
Here $\v S$ is a collective spin matrix, analogous to the collective spin vector $\vec S=\p{S_\x,S_\y,S_\z}$ in the case of SU(2) \cite{he2019engineering}.

We now discuss the spin Hamiltonian $H_{\t{int}}$ in Eq.~\eqref{eq:H_int}.
The operator $\v s_p\c\v s_q$ simply swaps the nuclear spin states of two atoms pinned to modes $p,q$.
The term $\v s_p\c\v s_q$ thereby assigns a definite energy of $+1$ ($-1$) to a pair of spins that are symmetric (anti-symmetric) under exchange.
In this sense, $\v s_p\c\v s_q$ is analogous to the enforcement of SU(2) spin alignment, which similarly assigns (distinct) definite energies to anti-symmetric spin-0 singlet $\ket{\up\dn}-\ket{\dn\up}$ and the symmetric spin-1 triplets $\set{\ket{\up\up},\ket{\dn\dn},\ket{\up\dn}+\ket{\dn\up}}$.

By summing over all pair-wise exchange terms $\v s_p\c\v s_q$, the interaction Hamiltonian $H_{\t{int}}$ energetically enforces a permutational symmetry among all spins, opening an energy gap $U$ between the manifold of all permutationally-symmetric (PS) states and the orthogonal complement of excited (e.g.~spin-wave) states that break permutational symmetry.
In the case of SU(2), the PS manifold is precisely the Dicke manifold of collective states $\ket{m_\z}$ with total spin $S=\frac{N}{2}$ and definite spin projection $m_\z\in\set{S,S-1,\cdots,-S}$ onto a fixed quantization axis.
Equivalently, Dicke states $\ket{m_\z}=\ket{m_\up,m_\dn}$ can be labeled by a definite number of spins $m_\up=S+m_\z$ ($m_\dn=S-m_\z$) pointing up (down) along the spin quantization axis, with $m_\up+m_\dn=N$.
In the general case of SU($n$), the PS manifold is similarly spanned by states $\ket{m_s,m_{s-1},\cdots,m_{-s}}$ with a definite number $m_\mu$ of spins in state $\mu$, and $\sum_\mu m_\mu=N$.
The dimension of the PS manifold, or the ground-state degeneracy of $H_{\t{int}}$, is equal to the number of ways of assigning $N$ identical spins to $n$ distinct internal states, or ${N+n-1 \choose n-1} \sim N^{n-1}$.

@@

Finally, we take a moment to discuss individual $n$-level spins.
Whenever possible, it is desirable to find geometric representations of abstract mathematical objects, such as the state of a quantum system.
Geometric representations facilitate the development of understanding and intuitions that can be difficult to extract from algebraic expressions.
For this reason, the state of a two-level spin, or a qubit, is commonly represented by a point on (or within) the Bloch sphere.
More generally, the state $\ket\psi$ of a $n$-level spin can be represented by a probability distribution\footnote{Strictly speaking, in order to make $Q_\psi$ (and $Q_\rho$) a probability distribution, one has to divide it by a normalization factor $\frac{4\pi}{d}$.
We omit this normalization factor for convenience.} $Q_\psi$ on the sphere $\SS^2$.
The value $Q_\psi\p{\v v}$ at a point $\v v\in\SS^2$ equal to the overlap of $\ket\psi$ with a pure state $\ket{\v v}$ maximally polarized along $\v v$: $Q_\psi\equiv\abs{\bk{\v v|\psi}}^2$ (see Figure \ref{fig:spin_dist}).
In the case of a mixed state $\rho$, this distribution is defined by $Q_\rho\equiv\bk{\v v|\rho|\v v}$.
The dimension $n$ is reflected in $Q_\rho$ by the fact that it can be decomposed into spherical harmonics $Y_{\ell m}$ with degree $\ell<n$ (that is, $Q_\rho$ is ``low resolution'' or ``band-limited'' at degree $n$) \cite{perlin2020qudit}, as well as in the normalization $\int_{\v v\in\SS^2}\dd^2\v v\,Q_\rho\p{\v v}=\frac{4\pi}{d}$.
Closely related spherical representations of multilevel spin operators are discussed in Ref.~\cite{li2013weylwignermoyal}.

@@

We wish to study the dynamical behaviors of experimentally realizable SU($n$) spin models.
In order to ensure validity of the $\v S\c\v S$ spin Hamiltonian in Eq.~\eqref{eq:H_int}, we consider initial preparation of a spin-polarized state\footnote{We remind the reader that validity of the spin model requires that each momentum mode is at most singly-occupied, which is guaranteed in a spin-polarized state due to Pauli blocking.}, which has the added benefit of being simple to prepare experimentally.
However, a spin polarized state is an eigenstate of both the kinetic and interaction spin Hamiltonians in Section \ref{sec:spin_model}, so we need additional ingredients for nontrivial behavior.
Here, we consider the addition of spin-orbit coupling (SOC) induced by external driving fields.

@@

The spin-$s$ generalization of Eq.~\eqref{eq:drive_2} is
\begin{align}
  H_{\t{drive}}^{(\phi)} = \Omega_0 \sum_{q,\mu} \bk{s,\mu;1,1|s,\mu+1}
  c_{q+\phi,\mu+1}^\dag c_{q\mu} + \t{h.c.},
\end{align}
where $\mu\in\set{s,s-1,\cdots,-s}$, and $\bk{s,\mu;1,1|s,\mu+1}$ is a Clebsch-Gordan coefficient.
The gauge gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ again takes $H_{\t{drive}}^{(\phi)}\to \Omega S_\x$ as in Eq.~\eqref{eq:drive_trans}, where now
\begin{align}
  \Omega = \f{\Omega_0}{\xi},
  &&
  \xi \equiv -\sqrt{\f{s(s+1)}{2}},
\end{align}
and
\begin{align}
  s_{\x,q} \equiv \xi \sum_\mu \bk{s,\mu;1,1|s,\mu+1} c_{q,\mu+1}^\dag c_{q\mu} + \t{h.c.}
  \label{eq:spin_x_op}
\end{align}
is an $n$-level spin-$x$ operator for spin $q$, with which $e^{-\i\theta s_{\x,q}}$ rotates the state of spin $q$ about the $x$ axis by an angle $\theta$.
Despite complicating details, the change from two-level to $n$-level spins thus results in a drive Hamiltonian with an identical physical interpretation: after moving into an appropriate ``gauged frame'', this drive simply rotates all spins about the $x$ axis at a rate $\Omega$.

\begin{figure}
\centering
\includegraphics{soc_panels.pdf}
\caption{
Spin-orbit coupling for 2-level ({\bf a},{\bf c}) and 4-level ({\bf b},{\bf d}) spins.
Colors indicate different spin projections $\mu$.
In the ``lab frame'' ({\bf a},{\bf b}), kinetic energy is insensitive to spin, but the drive imparts a momentum kick $q\to q+\phi$ into any atom that absorbs a photon.
Changing to the ``gauged frame'' ({\bf c},{\bf d}) makes the drive diagonal in momentum, but comes at the cost of making kinetic energy spin-dependent.
}
\label{fig:soc_panels}
\end{figure}

Of course, spin-orbit coupling cannot be ``gauged away'' entirely.
Making a gauge transformation to simplify the drive comes at the cost of making kinetic energy spin dependent, taking
\begin{align}
  H_{\t{kin}} \to H_{\t{kin}}^{(\phi)}
  \equiv -J \sum_q \cos\p{q+\mu\phi} s_{\mu\mu q},
\end{align}
as visualized in Figure \ref{fig:soc_panels}.
To better interpret this Hamiltonian, we can write it in the simplified form
\begin{align}
  H_{\t{kin}}^{(\phi)}
  = -J \sum_q
  \sp{\cos\p{q} w_{+,q}^{(\phi)} - \sin\p{q} w_{+,q}^{(\phi)}},
\end{align}
where
\begin{align}
  w_{+,q}^{(\phi)} &\equiv \sum_\mu \cos\p{\mu\phi} s_{\mu\mu q}, \\
  w_{-,q}^{(\phi)} &\equiv \sum_\mu \sin\p{\mu\phi} s_{\mu\mu q}.
\end{align}
For two-level spins, $w_{+,q}^{(\phi)}$ is proportional to the identity operator, and $w_{+,q}^{(\phi)}=2\sin\p{\phi/2} s_{q,\z}$, where $s_{\z,q}=\sum_\mu \mu\, s_{\mu\mu q}$ is a spin-$z$ operator for spin $q$, so the kinetic Hamiltonian in the gauged frame describes a (synthetic) inhomogeneous magnetic field:
\begin{align}
  \left. H_{\t{kin}}^{(\phi)} \right|_{n=2}
  = 2\sin\p{\phi/2} J \sum_q \sin\p{q} s_{\z,q}.
\end{align}
When $n>2$, an inhomogeneous magnetic field is likewise recovered in the limit of weak SOC, $s\phi\ll1$, in which case
\begin{align}
  \left. H_{\t{kin}}^{(\phi)} \right|_{s\phi\ll1}
  = J\phi \sum_q \sin\p{q} s_{\z,q} + O\p{(s\phi)^2}.
\end{align}
As the strength of SOC is increased, this Hamiltonian acquires terms with higher powers of $s_{\z,q}$, up to $s_{\z,q}^{n-1}$.

In principle, making the gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ also transforms the interaction Hamiltonian as
\begin{align}
  H_{\t{int}} \to H_{\t{int}}^{(\phi)}
  \equiv -\f{U}{2N} \sum_{p,q,\mu,\nu}
  c_{p-\mu\phi,\mu}^\dag c_{p-\nu\phi,\nu}
  c_{q-\nu\phi,\nu}^\dag c_{q-\mu\phi,\mu}.
\end{align}
However, $H_{\t{int}}^{(\phi)}$ cannot be exactly reduced to a spin Hamiltonian that addresses the same spin degrees of freedom as $H_{\t{kin}}^{(\phi)}$.
Nonetheless, for sufficiently weak SOC (small $\phi$) the effect of $H_{\t{int}}^{(\phi)}$ may be approximated by $H_{\t{int}}^{(0)}=H_{\t{int}}$, which has previously been benchmarked for SU(2)-symmetric interactions \cite{he2019engineering, smale2019observation}.
To ensure that $H_{\t{kin}}^{(\phi)}$ does not become trivial as $\phi\to0$, we can simultaneously take $J\to\infty$ while keeping $J\phi$ constant.
Altogether, the interacting spin Hamiltonian in the gauged frame becomes
\begin{align}
  H_{\t{spin}} = -\f{U}{2N} \v S\c\v S + J\phi \sum_q \sin\p{q} s_{\z,q}.
  \label{eq:H_spin}
\end{align}

@@

As shown in insets of Figure \ref{fig:mean_op_ss_X}, mean-field results for different spin dimensions $n$ collapse onto each other when normalizing the field $J\phi/U$ to its critical value, $\p{J\phi/U}\to\p{J\phi/U}\times\p{n/2}^{1/3}$, and rescaling $\bbk{\bar\sigma_x}_\MF\to\bbk{\bar\sigma_x}_\MF/\gamma\p{n/2}$ and $\bbk{\ss}_\MF\to\bbk{\ss}_\MF/\gamma\p{n}-1$, where
\begin{align}
  \gamma\p{k} \equiv \f{\Gamma\p{k-\frac12}}{\sqrt\pi\,\Gamma\p{k}}.
  \label{eq:gamma}
\end{align}
The rescaling of transverse magnetization and interaction energy can be understood by considering their limiting behavior as $J\phi/U\to\infty$ or $J\phi/U\to0^+$.

>>>>

% more shorthands
\newcommand{\up}{\uparrow}
\newcommand{\dn}{\downarrow}
\newcommand{\x}{\text{x}}
\newcommand{\y}{\text{y}}
\newcommand{\z}{\text{z}}
\newcommand{\g}{\text{g}}
\newcommand{\e}{\text{e}}
\renewcommand{\SS}{\mathbb{S}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\P}{\mathcal{P}}

\newcommand{\su}{\mathfrak{su}}
\newcommand{\MF}{\text{MF}}
\renewcommand{\ss}{\bar{\v s}\c\bar{\v s}}
\DeclareMathOperator{\diag}{diag}

@@

Here we review the realization of a collective SU($n$) spin model in a system of ultracold alkaline-earth(-like) atoms on an optical lattice.
Without external driving fields, the evolution of such atoms in their electronic ground state is governed by single-body kinetic and two-body interaction Hamiltonians of the form
\begin{align}
  H_{\t{kin}}
  &= -\f{J}{2} \sum_{\bk{j,j'},\mu} c_{j\mu}^\dag c_{j'\mu} + \t{h.c.}, \\
  H_{\t{int}}
  &= \f{U_0}{2} \sum_{j,\mu,\nu}
  c_{j\mu}^\dag c_{j\mu} c_{j\nu}^\dag c_{j\nu},
\end{align}
where $\bk{j,j'}$ denotes neighboring lattice sites $j$ and $j'$; $\mu,\nu\in\set{s,s-1,\cdots,-s}$ index orthogonal spin states of a spin-$s$ nucleus, with $s=\frac{n-1}{2}$ (e.g.~$s=\frac{9}{2}$ in the case of ${}^{87}$Sr with $10$ nuclear spin states); $c_{j\mu}$ is a fermionic annihilation operator, $J$ is a tunneling amplitude; and $U_0$ is a two-body on-site interaction energy.
In the present work, we neglect inter-site interactions and interaction-assisted hopping, which may become relevant in a sufficiently shallow lattice, namely when $J\gtrsim E_{\t{R}}$, where $E_{\t{R}}$ is the atomic lattice recoil energy.
For simplicity, we now assume a periodic lattice of $L$ sites, and expand the on-site fermionic operators in terms of operators addressing (quasi-)momentum modes $q$, $c_{j\mu}=\frac1{\sqrt{L}}\sum_k e^{-\i q\c j} c_{q\mu}$, finding that
\begin{align}
  H_{\t{kin}}
  &= -J\sum_{q,\mu} \cos\p{q} c_{q\mu}^\dag c_{q\mu}, \\
  H_{\t{int}}
  &= \f{U}{2N} \sum_{k,\ell,p,q,\mu,\nu}
  c_{k\mu}^\dag c_{\ell\mu} c_{p\nu}^\dag c_{q\nu}
  \times \delta_{k+p,\ell+q},
  \label{eq:H_int_momenta}
\end{align}
where $N$ is the total number of atoms on the lattice, we define $U\equiv U_0\times N/L$ for convenience, $\delta_{k+p,\ell+q}=1$ if $k+p=\ell+q$ and zero otherwise (enforcing conservation of momentum), and we work in units with lattice spacing $a=1$.

If the interaction energy $U_0$ is smaller than the single-particle bandwidth $\sim J$, then the mode-changing collisions in $H_{\t{int}}$ become off-resonant, motivating the frozen-mode approximation $\set{k,p}=\set{\ell,q}$ (which we assume for the remainder of this work).
The operator content of terms with $k=\ell$ and $p=q$ takes the form $c_{\ell\mu}^\dag c_{\ell\mu} c_{q\nu}^\dag c_{q\nu}$, so these terms contribute an overall energy $\frac12NU$ that can be neglected in the absence of coherent atom number fluctuations.
Defining the spin operators $s_{\mu\nu q}\equiv c_{q\mu}^\dag c_{q\nu}$, the remaining terms of the kinetic and interaction Hamiltonians become
\begin{align}
  H_{\t{kin}} &= -J\sum_{q,\mu} \cos\p{q} s_{\mu\mu q},
  \label{H_kin_start} \\
  H_{\t{int}} &= -\f{U}{2N} \sum_{p,q,\mu,\nu} s_{\mu\nu p} s_{\nu\mu q}.
  \label{eq:H_int_start}
\end{align}
Throughout this work, we will assume that atomic modes are singly-occupied, e.g.~due to the initialization of a spin-polarized state with one atom per lattice site, in which multiple occupation of an atomic mode is forbidden by fermionic statistics (Pauli exclusion).
In this case, we can neglect the fermionic origins of $H_{\t{int}}$ and instead treat our system as a bona fide spin model.
Note that the ``kinetic'' terms of this spin model ($H_{\t{kin}}$) are proportional to the identity operator, and can be neglected for the time being.
The fermionic origin of these terms will return to play an important role when we consider spin-orbit coupling in Section \ref{sec:SOC}.

To further simplify the interaction Hamiltonian $H_{\t{int}}$ and write it in a form suggestively similar to SU(2) spin alignment, we now construct the operator-valued spin matrix
\begin{align}
  \v s_q \equiv \sum_{\mu,\nu} s_{\mu\nu q} \op{\mu}{\nu},
\end{align}
and for any pair of such operator-valued matrices $\v A,\v B$, we define the inner product
\begin{align}
  \v A \c \v B \equiv \Tr_{\t{aux}}\p{\v A^\dag\v B}
  = \sum_{\mu,\nu} A_{\mu\nu}^\dag B_{\mu\nu},
\end{align}
where the trace is performed over the auxiliary matrix degrees of freedom.
These definitions allow us to write the spin Hamiltonian in Eq.~\eqref{eq:H_int_start} as
\begin{align}
  H_{\t{int}} = -\f{U}{2N} \sum_{p,q} \v s_p\c\v s_q
  = -\f{U}{2N}\v S\c\v S,
  &&
  \v S \equiv \sum_q \v s_q.
  \label{eq:H_int}
\end{align}
Here $\v S$ is a collective spin matrix, analogous to the collective spin vector $\vec S=\p{S_\x,S_\y,S_\z}$ in the case of SU(2) \cite{he2019engineering}, with $\v S\c\v S \simeq 2 \vec S\c\vec S = 2 \sum_{\alpha\in\set{\x,\y,\z}}S_\alpha^2$ when $n=2$ (with $\simeq$ denoting equality after modding out identity terms).

We now discuss the spin Hamiltonian $H_{\t{int}}$ in Eq.~\eqref{eq:H_int}.
The operator $\v s_p\c\v s_q$ simply swaps the nuclear spin states of two atoms pinned to modes $p,q$.
The term $-\v s_p\c\v s_q$ thereby assigns a definite energy of $-1$ ($+1$) to a pair of spins that are symmetric (anti-symmetric) under exchange.
In this sense, $\v s_p\c\v s_q$ is analogous to the enforcement of SU(2) spin alignment by ferromagnetic interactions, which similarly assigns (distinct) definite energies to anti-symmetric spin-0 singlet $\ket{\up\dn}-\ket{\dn\up}$ and the symmetric spin-1 triplets $\set{\ket{\up\up},\ket{\dn\dn},\ket{\up\dn}+\ket{\dn\up}}$.

By summing over all pair-wise exchange terms $\v s_p\c\v s_q$, the interaction Hamiltonian $H_{\t{int}}$ energetically enforces a permutational symmetry among all spins, opening an energy gap $U$ between the manifold of all permutationally-symmetric (PS) states and the orthogonal complement of excited (e.g.~spin-wave) states that break permutational symmetry.
In the case of SU(2), the PS manifold is precisely the Dicke manifold of collective states $\ket{m_\z}$ with total spin $S=\frac{N}{2}$ and definite spin projection $m_\z\in\set{S,S-1,\cdots,-S}$ onto a fixed quantization axis.
Equivalently, Dicke states $\ket{m_\z}=\ket{m_\up,m_\dn}$ can be labeled by a definite number of spins $m_\up=S+m_\z$ ($m_\dn=S-m_\z$) pointing up (down) along the spin quantization axis, with $m_\up+m_\dn=N$.
In the general case of SU($n$), the PS manifold is similarly spanned by states $\ket{m_s,m_{s-1},\cdots,m_{-s}}$ with a definite number $m_\mu$ of spins in state $\mu$, and $\sum_\mu m_\mu=N$.
The dimension of the PS manifold is equal to the number of ways of assigning $N$ identical spins to $n$ distinct internal states, or ${N+n-1 \choose n-1} \sim N^{n-1}$.

External fields or additional interactions that respect permutational symmetry can induce nontrivial dynamics within the PS manifold.
Even if additional Hamiltonian terms break permutational symmetry, as long as the coupling to states outside of the PS manifold is small (compared to the energy gap $U$) these terms can be restricted to the PS manifold perturbatively (see Appendix \ref{sec:pert_theory}).
Simulating dynamics within the PS manifold requires calculating matrix elements $\bk{\ell|\O|m}$ of spin operators $\O$ with respect to PS states $\ket\ell,\ket{m}$; we discuss this calculation in Appendix \ref{sec:PS_ops}.

@@

Finally, we take a moment to discuss individual $n$-level spins.
Whenever possible, it is desirable to find geometric representations of abstract mathematical objects, such as the state of a quantum system.
Geometric representations facilitate the development of understanding and intuitions that can be difficult to extract from algebraic expressions alone.
For this reason, the state of a two-level spin, or a qubit, is commonly represented by a point on (or within) the Bloch sphere.
More generally, the state $\ket\psi$ of a $n$-level spin can be represented by a probability distribution\footnote{Strictly speaking, in order to make $Q_\psi$ (and $Q_\rho$) a probability distribution, one has to divide it by a normalization factor $\frac{4\pi}{d}$.
We omit this normalization factor for convenience.} $Q_\psi$ on the sphere $\SS^2$.
The value $Q_\psi\p{\v v}$ at a point $\v v\in\SS^2$ equal to the overlap of $\ket\psi$ with a pure state $\ket{\v v}$ maximally polarized along $\v v$: $Q_\psi\equiv\abs{\bk{\v v|\psi}}^2$ (see Figure \ref{fig:spin_dist}).
In the case of a mixed state $\rho$, this distribution is defined by $Q_\rho\equiv\bk{\v v|\rho|\v v}$.
The dimension $n$ is reflected in $Q_\rho$ by the fact that it can be decomposed into spherical harmonics $Y_{\ell m}$ with degree $\ell<n$ (that is, $Q_\rho$ is ``low resolution'' or ``band-limited'' at degree $n$) \cite{perlin2020qudit}, as well as in the normalization $\int_{\v v\in\SS^2}\dd^2\v v\,Q_\rho\p{\v v}=\frac{4\pi}{d}$.
Closely related spherical representations of multilevel spin states (and operators) are discussed in Ref.~\cite{li2013weylwignermoyal}.

@@

We wish to study the dynamical behaviors of experimentally realizable SU($n$) spin models.
In order to ensure validity of the $\v S\c\v S$ spin Hamiltonian in Eq.~\eqref{eq:H_int}, we consider initial preparation of a spin-polarized state\footnote{We remind the reader that validity of the spin model requires that each momentum mode is at most singly-occupied, which is guaranteed for a spin-polarized state due to Pauli blocking.}, which has the added benefit of being simple to prepare experimentally.
However, a spin polarized state is an eigenstate of the $\v S\c\v S$ spin Hamiltonian, so we need additional ingredients for nontrivial behavior.
Here, we consider the addition of spin-orbit coupling (SOC) induced by external driving fields.

@@

The spin-$s$ generalization of Eq.~\eqref{eq:drive_2} is
\begin{align}
  H_{\t{drive}}^{(\phi)} = \Omega_0 \sum_{q,\mu} \bk{s,\mu;1,1|s,\mu+1}
  c_{q+\phi,\mu+1}^\dag c_{q\mu} + \t{h.c.},
\end{align}
where $\mu\in\set{s,s-1,\cdots,-s}$, and $\bk{s,\mu;1,1|s,\mu+1}$ is a Clebsch-Gordan coefficient.
The gauge gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ again takes $H_{\t{drive}}^{(\phi)}\to \Omega S_\x$ as in Eq.~\eqref{eq:drive_trans}, where now
\begin{align}
  \Omega = \f{\Omega_0}{\xi},
  &&
  \xi \equiv -\sqrt{\f{s(s+1)}{2}},
\end{align}
and
\begin{align}
  s_{\x,q} \equiv \xi \sum_\mu \bk{s,\mu;1,1|s,\mu+1} c_{q,\mu+1}^\dag c_{q\mu} + \t{h.c.}
  \label{eq:spin_x_op}
\end{align}
is an $n$-level spin-$x$ operator for spin $q$, with which $e^{-\i\theta s_{\x,q}}$ rotates the state of spin $q$ about the $x$ axis by an angle $\theta$ (literally rotating the sphere in Figure \ref{fig:spin_dist}).
Despite complicating details, the change from two-level to $n$-level spins thus results in a drive Hamiltonian with an identical physical interpretation: after moving into an appropriate ``gauged frame'', this drive simply rotates all spins about the $x$ axis at a rate $\Omega$.

\begin{figure}
\centering
\includegraphics{soc_panels.pdf}
\caption{
Spin-orbit coupling for 2-level ({\bf a},{\bf c}) and 4-level ({\bf b},{\bf d}) spins.
Colors indicate different spin projections $\mu$.
In the ``lab frame'' ({\bf a},{\bf b}), kinetic energy is insensitive to spin, but the drive imparts a momentum kick $q\to q+\phi$ into any atom that absorbs a photon.
Changing to the ``gauged frame'' ({\bf c},{\bf d}), essentially by shifting the momentum label $q$ for each spin state $\mu$, makes the drive diagonal in the momentum index, but comes at the cost of making kinetic energy spin-dependent.
}
\label{fig:soc_panels}
\end{figure}

Of course, spin-orbit coupling cannot be ``gauged away'' entirely.
Making a gauge transformation to simplify the drive comes at the cost of making kinetic energy spin-dependent, taking
\begin{align}
  H_{\t{kin}} \to H_{\t{kin}}^{(\phi)}
  \equiv -J \sum_q \cos\p{q+\mu\phi} s_{\mu\mu q},
\end{align}
as visualized in Figure \ref{fig:soc_panels}.
To better interpret this Hamiltonian, we can write it in the simplified form
\begin{align}
  H_{\t{kin}}^{(\phi)}
  = -J \sum_q
  \sp{\cos\p{q} w_{+,q}^{(\phi)} - \sin\p{q} w_{+,q}^{(\phi)}},
\end{align}
where
\begin{align}
  w_{+,q}^{(\phi)} &\equiv \sum_\mu \cos\p{\mu\phi} s_{\mu\mu q}, \\
  w_{-,q}^{(\phi)} &\equiv \sum_\mu \sin\p{\mu\phi} s_{\mu\mu q}.
\end{align}
For two-level spins, $w_{+,q}^{(\phi)}$ is proportional to the identity operator, and $w_{+,q}^{(\phi)}=2\sin\p{\phi/2} s_{q,\z}$, where $s_{\z,q}=\sum_\mu \mu\, s_{\mu\mu q}$ is a spin-$z$ operator for spin $q$, so the kinetic Hamiltonian in the gauged frame describes a (synthetic) inhomogeneous magnetic field:
\begin{align}
  \left. H_{\t{kin}}^{(\phi)} \right|_{n=2}
  = 2\sin\p{\phi/2} J \sum_q \sin\p{q} s_{\z,q}.
\end{align}
When $n>2$, an inhomogeneous magnetic field is likewise recovered in the limit of weak SOC, $s\phi\ll1$, in which case
\begin{align}
  \left. H_{\t{kin}}^{(\phi)} \right|_{s\phi\ll1}
  = J\phi \sum_q \sin\p{q} s_{\z,q} + O\p{(s\phi)^2}.
\end{align}
As the strength of SOC is increased, this Hamiltonian acquires terms with higher powers of $s_{\z,q}$, up to $s_{\z,q}^{n-1}$.

In principle, making the gauge transformation $c_{q\mu}^\dag\to c_{q-\mu\phi,\mu}^\dag$ also transforms the interaction Hamiltonian as
\begin{align}
  H_{\t{int}} \to H_{\t{int}}^{(\phi)}
  \equiv -\f{U}{2N} \sum_{p,q,\mu,\nu}
  c_{p-\mu\phi,\mu}^\dag c_{p-\nu\phi,\nu}
  c_{q-\nu\phi,\nu}^\dag c_{q-\mu\phi,\mu}.
\end{align}
However, $H_{\t{int}}^{(\phi)}$ cannot be exactly reduced to a spin Hamiltonian that addresses the same spin degrees of freedom as $H_{\t{kin}}^{(\phi)}$.
Nonetheless, for sufficiently weak SOC (small $s\phi$) the effect of $H_{\t{int}}^{(\phi)}$ may be approximated by $H_{\t{int}}^{(0)}=H_{\t{int}}$, which has previously been benchmarked for SU(2)-symmetric interactions \cite{he2019engineering, smale2019observation}.
To ensure that $H_{\t{kin}}^{(\phi)}$ does not become trivial as $\phi\to0$, we can simultaneously take $J\to\infty$ ($J$ large compared to $U$) while keeping $J\phi$ constant.
Altogether, the interacting spin Hamiltonian in the gauged frame becomes
\begin{align}
  H_{\t{spin}} = -\f{U}{2N} \v S\c\v S + J\phi \sum_q \sin\p{q} s_{\z,q}.
  \label{eq:H_spin}
\end{align}

@@

As shown in insets of Figure \ref{fig:mean_op_ss_X}, mean-field results for different spin dimensions $n$ collapse onto each other when normalizing the field $J\phi/U$ to its critical value, $\p{J\phi/U}\to\p{J\phi/U}\times\p{n/2}^{1/3}$, and rescaling
\begin{align}
  \bbk{\bar\sigma_x}_\MF \to \f{\bbk{\bar\sigma_x}_\MF}{\gamma\p{n/2}},
  &&
  \bbk{\ss}_\MF \to \f{\bbk{\ss}_\MF-\gamma\p{n}}{1-\gamma\p{n}},
  \label{eq:rescale}
\end{align}
where
\begin{align}
  \gamma\p{k} \equiv \f{\Gamma\p{k-\frac12}}{\sqrt\pi\,\Gamma\p{k}}
  \stackrel{k\ge2}{\approx} \f1{\sqrt{\pi(k-1)}}.
  \label{eq:gamma}
\end{align}
The rescaling of transverse magnetization and interaction energy can be understood by considering their limiting behavior as $J\phi/U\to\infty$ or $J\phi/U\to0^+$.
